---
title: "SPOTIFY ATTRIBUTES REGRESSION PROJECT"
author: "Kaustubh Dangche, Lakshmi Sindhu Pulugundla & Lasya Priya Thota"
date: "2024-11-08"
output:
  html_document:
    output_dir: ./outputs/Spotify-Regression-Project  # Specifies the outputs folder for HTML
    df_print: paged
  pdf_document:
    output_dir: ./outputs/Spotify-Regression-Project  # Specifies the outputs folder for PDF
    latex_engine: xelatex
---


```{r setup, include=FALSE}
# Set the working directory to the directory where the .Rmd file is located
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Check if outputs folder exists, if not create it
if (!dir.exists("./outputs/Spotify-Regression-Project")) {
  dir.create("./outputs/Spotify-Regression-Project", recursive = TRUE)
}

# knitr settings
knitr::opts_chunk$set(echo = TRUE)

```


# **Project Objective and Overview**

The objective of this project is to predict the **`msPlayed`** (playtime in milliseconds) of songs on **Spotify** based on various song attributes. These attributes include both **continuous** variables such as **danceability**, **energy**, **loudness**, and **tempo**, as well as **categorical** features like **genre** and **artistName**.

The project’s core goal is to understand how different musical attributes influence the total playtime of songs and build a regression model to predict future playtime. The dataset includes a variety of song characteristics that provide insights into musical features, which can be useful for music industry professionals and data scientists alike.

By the end of this project, the following objectives will be achieved:
- Build a robust **regression model** to predict **`msPlayed`**.
- Understand the significance of various musical features in predicting playtime.
- Clean and preprocess the data, handling missing values, duplicates, and categorical features.
- Perform **Exploratory Data Analysis (EDA)** to visualize trends and patterns in the data.

---

### **Focus of the Project**:

- **Primary Application**: **Music Playtime Prediction**
  - Predicting the playtime (`msPlayed`) based on the attributes of the song.
  
- **Core Model**: **Regression Model (Multiple Linear Regression / Polynomial Regression)**
  - Using regression analysis to model the relationship between song attributes and playtime.

---

### **Phase-Wise Project Flow**

#### **Phase 1: Data Preprocessing & Basic Data Understanding**

1. **Data Loading and Initial Exploration**  
   - Loading the Spotify Song Attributes dataset and inspecting its structure.
   
2. **Data Cleaning and Type Modification**  
   - Converting columns into appropriate formats and removing redundant or irrelevant columns.
   
3. **Handling Missing Values and Checking for Duplicates**  
   - Ensuring no missing values or duplicate records exist.
   
4. **Outlier Detection and Handling**  
   - Identifying and addressing any extreme outliers in key features.

5. **Feature Scaling**  
   - Standardizing and normalizing continuous features, making them suitable for regression modeling.

---

#### **Phase 2: Exploratory Data Analysis (EDA)**

1. **Visualizing Distribution of Attributes**  
   - Visualizing how each feature, especially continuous variables like `danceability`, `energy`, and `msPlayed`, is distributed.

2. **Exploring Relationships Between Features**  
   - Using scatter plots, correlation matrices, and other visual tools to explore relationships between the features and `msPlayed`.

---

#### **Phase 3: Feature Engineering & Extraction**

1. **Handling Categorical Data**  
   - Converting categorical variables like `genre` and `artistName` into factors, and preparing them for model fitting.

2. **Creating New Features**  
   - Deriving new features from the existing ones to enhance the predictive power of the model.

---

#### **Phase 4: Model Building**

1. **Model Selection (Linear/Multiple Regression)**  
   - Selecting a suitable regression model to predict `msPlayed`.

2. **Model Evaluation**  
   - Evaluating the model performance using metrics such as R-squared, RMSE, and residual plots.

---

#### **Phase 5: Post-Modeling Analysis**

1. **Model Refinement**  
   - Making improvements to the model based on diagnostic tests and evaluation metrics.

---

#### **Phase 6: Conclusion and Future Enhancements**

1. **Summarizing Findings**  
   - Discussing the effectiveness of the regression model and the insights gained.

2. **Future Work**  
   - Suggesting improvements, such as trying more complex models (e.g., Random Forest, Gradient Boosting), or exploring additional features for better predictions.

---

# **1. Data Collection and Pre-Processing**

## **1.1 Load the Necessary Libraries & Dataset**

```{r}
# Load necessary libraries
library(dplyr)  # For data manipulation
library(ggplot2)  # Optional, for visualization if needed

# Load the Spotify attribute dataset
data <- read.csv("./data/Spotify_Song_Attributes.csv")
# Create a copy of the spotify attribute dataset
spotify_data <- data
# View the first few rows to understand the structure
head(spotify_data)
```

### **Inference**:
The dataset is successfully loaded into R, and a copy (`spotify_data`) is created to preserve the original data. We can now inspect the first few rows to understand its structure.

## **1.2 View Dataset Structure**

```{r}
# View the structure of the data
str(spotify_data)
```

### **Inference**:
The `str()` function provides a quick look at the internal structure of the dataset. We can check the data types (e.g., numeric, factor) of each column, which helps in deciding how to clean or transform the data.

## **1.3 Summary of Each Column**

```{r}
# View a summary of each column (e.g., min, max, mean, etc.)
summary(spotify_data)
```

### **Inference**:
The summary statistics allow us to understand the distribution of each feature, including mean, min, max, and standard deviation. This is crucial for identifying outliers and deciding which variables require special treatment.

## **1.4 Handling Missing Values**

```{r}
# Check for missing values in each column
colSums(is.na(spotify_data))
```

#### **We can see**:
1. **Columns with Zero Missing Values**: Columns like `trackName`, `artistName`, `msPlayed`, `genre`, `id`, `uri`, `track_href`, `analysis_url`, and `type` have no missing values, ensuring they are complete for analysis.
2. **Columns with 550 Missing Values**: Columns such as `danceability`, `energy`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `duration_ms`, and `time_signature` have significant missing values (550). This indicates a need to address these missing entries to ensure reliable analysis.

### **1.4.1 Remove Missing Values**

```{r}
# Option 1: Remove rows with missing values (if they are few)
spotify_data <- na.omit(spotify_data)
```

### **Inference**:
Using `na.omit()` removes rows with missing data, leaving us with only complete observations. This is necessary for regression models, which cannot handle missing values.

### **1.4.2 Check Missing Values After Removal**

```{r}
# Check for missing values in each column after applying na.omit function 
colSums(is.na(spotify_data))
```

#### **All values being zero in the output confirms** that no missing data remains in the dataset. This clean dataset now contains only complete rows, ideal for accurate and consistent analysis.

### **Just check missing values removed from structure data or not.**

```{r}
str(spotify_data)
```

## **1.5 Remove Unwanted Columns**

```{r}
# Load dplyr library
library(dplyr)

# Remove `trackName` along with other unnecessary columns
spotify_data <- spotify_data %>% select(-id, -uri, -track_href, -analysis_url, -trackName, -type)
```

### **Statistical Reasoning for Dropping Variables:**
- **`id`, `uri`, `track_href`, `analysis_url`, `trackName`, `type`**: These columns do not contribute to predicting `msPlayed`. They are either unique identifiers or non-informative for the analysis.

```{r}
# Check the specified columns have been removed or not
str(spotify_data)
```

## **1.6 Data Cleaning**

### **1.6.1 Check Whitespace in `artistName` Column**

```{r}
# Check for leading or trailing whitespace
whitespace_issues <- spotify_data$artistName[grepl("^\\s|\\s$", spotify_data$artistName)]

# Display entries with whitespace issues
print(whitespace_issues)
```

### **Inference**:
Whitespace in `artistName` can cause inconsistencies, so it is necessary to clean these entries.

### **1.6.2 Identify Special Characters in `artistName`**

```{r}
# Identify special characters in artist names
special_characters <- spotify_data$artistName[grepl("[^a-zA-Z0-9\\s]", spotify_data$artistName)]

# Display total number of entries with special characters and the first 25 entries
cat("Total number of ntries with Special Characters:", length(special_characters), "\n")
head(special_characters, 25)
```

### **Inference**:
Special characters in `artistName` can cause inconsistencies when grouping artists. Identifying these characters helps ensure that they are cleaned properly.

### **1.6.3 Identifying Duplicated Artist Names**

```{r}
# Identify duplicated artist names after cleaning
duplicate_artists <- spotify_data$artistName[duplicated(spotify_data$artistName)]

# Display total number of duplicates and the first 10 duplicate entries
cat("Total number of duplicates:", length(duplicate_artists), "\n")
head(duplicate_artists, 10)
```

### **Inference**:
This output shows the total number of duplicate artist names and the first 10 duplicates, helping confirm that duplicate names are properly handled.

### **1.6.4 Replace Special Characters**

```{r}
# Replace special characters like $ with s (only if necessary)
spotify_data$artistName <- gsub("\\$", "s", spotify_data$artistName)
```

### **Inference**:
Replacing special characters ensures that `artistName` values are consistent, removing unwanted characters that may interfere with analysis.

### **1.6.5 Convert Artist Names to Lowercase**

```{r}
# Convert artistName to lowercase
spotify_data$artistName <- tolower(spotify_data$artistName)
```

### **Inference**:
Lowercasing ensures that variations in capitalization are removed, making artist names uniform for analysis.

### **1.6.6 Removing Identical Rows**

```{r}
# Removing rows that are identical across all columns
spotify_data <- distinct(spotify_data)
```

### **Inference**:
Removing identical rows ensures that only unique entries are retained, reducing redundancy in the dataset.

### **1.6.7 Group Rare Artists into "Other"**

```{r}
# Group artists with fewer than the threshold of entries into "Other"
threshold <- 5
artist_counts <- table(spotify_data$artistName)
rare_artists <- names(artist_counts[artist_counts < threshold])
spotify_data$artistName <- ifelse(spotify_data$artistName %in% rare_artists, "Other", spotify_data$artistName)
```

### **Inference**:
Grouping rare artists under the "Other" category reduces the number of unique categories, simplifying the model and making it easier to interpret.

### **1.6.8 Convert Artist Names to Factor**

```{r}
# Convert artistName to a factor
spotify_data$artistName <- as.factor(spotify_data$artistName)
```

### **Inference**:
Converting `artistName` to a factor ensures that it is correctly handled as a categorical variable in regression modeling.

---

### **1.6.9 Let's Clean the Genre Column**

```{r}
# Checking the genre data
head(spotify_data$genre)

head(unique(spotify_data$genre), 25)
```

```{r}
# Count missing values
sum(is.na(spotify_data$genre) | spotify_data$genre == "")
```

#### **Output Interpretation**:  
This output helps identify how many `genre` values are missing or empty, guiding further cleaning steps.

```{r}
# Replace empty genre values with "unknown"
spotify_data$genre[spotify_data$genre == ""] <- "unknown"
```

#### **Inference**:
Replacing empty genre values with "unknown" ensures that no missing data is mistakenly treated as an empty string.

```{r}
# Convert genre names to lowercase
spotify_data$genre <- tolower(spotify_data$genre)
```

```{r}
# Display the frequency of genres
genre_counts <- sort(table(spotify_data$genre), decreasing = TRUE)
print(genre_counts)
```

#### **Inference**:
The frequency table provides insights into the distribution of genres, helping identify the most common and rare genres for possible standardization.

```{r}
# Replace empty strings and unknown genres with NA
spotify_data$genre[spotify_data$genre == "" | spotify_data$genre == "unknown"] <- NA
```

```{r}
# Define a more comprehensive mapping for genre standardization
genre_mapping <- c(
    # Standardizing popular genres
    "alt z" = "alternative", 
    "album rock" = "rock", 
    "british orchestra" = "classical", 
    "desi hip hop" = "hip hop", 
    "bedroom r&b" = "r&b", 
    "singer-songwriter pop" = "pop", 
    "la pop" = "pop", 
    "lo-fi chill" = "lo-fi", 
    "orchestral soundtrack" = "classical", 
    "comic" = "other", 
    "alternative metal" = "metal", 
    "deep underground hip hop" = "hip hop", 
    "pop" = "pop", 
    "classical" = "classical", 
    "modern alternative rock" = "alternative", 
    "scandipop" = "pop", 
    "punjabi pop" = "pop", 
    "folk-pop" = "folk", 
    "acoustic pop" = "pop", 
    "art pop" = "pop", 
    "electronica" = "electronic", 
    "dance pop" = "pop", 
    "bedroom pop" = "pop", 
    "chill r&b" = "r&b", 
    "indian lo-fi" = "lo-fi", 
    "instrumental post-rock" = "post-rock", 
    "classic bollywood" = "bollywood", 
    "afghan pop" = "pop", 
    "classic rock" = "rock", 
    "german soundtrack" = "classical", 
    "anime lo-fi" = "lo-fi", 
    "lo-fi study" = "lo-fi", 
    "dark r&b" = "r&b", 
    "modern indie pop" = "indie", 
    "pop edm" = "edm", 
    "uk contemporary r&b" = "r&b", 
    "emo rap" = "hip hop", 
    "classic pakistani pop" = "pop", 
    "japanese chillhop" = "chillhop", 
    "japanese vgm" = "vgm", 
    "anime" = "other", 
    "bhangra" = "indian", 
    "afrobeats" = "afrobeat", 
    "j-pop" = "asian pop", 
    "k-pop" = "asian pop", 
    
    # More specific mappings for sub-genres
    "aggressive phonk" = "phonk",
    "alabama indie" = "indie",
    "ambient" = "ambient",
    "alabama rap" = "rap",
    "australian dance" = "dance",
    "australian indie" = "indie",
    "australian pop" = "pop",
    "austindie" = "indie",
    "bass trap" = "trap",
    "bedroom soul" = "soul",
    "big room" = "edm",
    "brostep" = "brostep",
    "calming instrumental" = "instrumental",
    "chillhop" = "chillhop",
    "chillstep" = "chillstep",
    "complextro" = "electronic",
    "contemporary country" = "country",
    "country" = "country",
    "deep tropical house" = "edm",
    "desi pop" = "pop",
    "detroit hip hop" = "hip hop",
    "detroit indie" = "indie",
    "disco" = "disco",
    "electropop" = "electropop",
    "electro house" = "edm",
    "electronic" = "electronic",
    "filmi" = "bollywood",
    "future rock" = "rock",
    "indietronica" = "indie",
    "indie pop rap" = "indie rap",
    "instrumental grime" = "grime",
    "instrumental math rock" = "post-rock",
    "j-pop boy group" = "j-pop",
    "japanese old school hip hop" = "hip hop",
    "k-pop girl group" = "k-pop",
    "lo-fi brasileiro" = "lo-fi",
    "lo-fi indie" = "lo-fi",
    "lo-fi jazzhop" = "lo-fi",
    "lo-fi latino" = "lo-fi",
    "melodic dubstep" = "dubstep",
    "metropopolis" = "electronic",
    "phonk" = "phonk",
    "pop edm" = "edm",
    "pop folk" = "folk",
    "rap rock" = "rap",
    "reggaeton" = "reggaeton",
    "soul" = "soul",
    "trap" = "trap",
    "vaporwave" = "vaporwave"
)

# Apply recode function for genre standardization
spotify_data$genre <- recode(spotify_data$genre, !!!genre_mapping)

# Final verification of genres
unique_genres <- unique(spotify_data$genre)
print(unique_genres)

```

#### **Inference**:
This code standardizes the genre column using the mapping defined in `genre_mapping`. The genres are recoded into broader categories, and the final unique values are verified.

#### Finally done with both categorical variables. So, I have saved this file in my directory :-  
 
```{r}
# Save the modified dataframe to a new CSV file
write.csv(spotify_data, "./data/spotify_dataset_modified.csv", row.names = FALSE)

```
